---
title: "HW4"
output:
  pdf_document: default
  html_document: default
---

```{r}
periodogram <- function(x){
  xPer <- (1/length(x))*abs(fft(x)^2)
  f <- seq(0,1.0-1/length(x),by=1/length(x))
  return(list(f,xPer))
}
```


\section{1}
$$Z_t \sim WN(1)$$
\subsection{a}
$$X_t = X_{t-1} - 0.9X_{t-2} + Z_t$$
```{r}
library(astsa)
freq = arma.spec(ar = c(1,-0.9), var.noise = 1, type = "line")
best = freq$freq[which.max(freq$spec)]
```
The frequencies that appear to be dominant are $f \in [0.125, 0.175]$. The peak occured at `r best` 

```{r}
set.seed(1)
sim = arima.sim(n = 50, list(ar = c(1,-0.9)))
period = cumsum(rle(as.vector(sign(diff(sim))))$lengths) + 1
plot(sim)
abline(v = period, col = 'red')
```
```{r}
plot(periodogram(sim)[[1]],
     periodogram(sim)[[2]], type = 'l',
     xlim = c(0,0.5))
```


```{r}
mean(rle(as.vector(sign(diff(sim))))$lengths) * 2
1/best
```
It appears that the observed period matches reasonably well with $\frac{1}{f}$

\newpage
\subsection{b}
$$X'_t = -0.3X'_{t-2} - 0.9X'_{t-4} + Z_t$$
```{r}
library(astsa)
freq = arma.spec(ar = c(0,-0.3,0,-0.9), var.noise = 1, type = "line")
best1 = freq$freq[which.max(freq$spec[1:300])]
best2 = freq$freq[which.max(freq$spec)]
```
The frequencies that appear to be dominant are $f_1 \in [0.125, 0.15]$ and $f_1 \in [0.35, 0.375]$. The peaks occured at `r best1` and `r best2` 

```{r}
set.seed(1)
sim = arima.sim(n = 50, list(ar = c(0,-0.3,0,-0.9)))
period = cumsum(rle(as.vector(sign(diff(sim))))$lengths) + 1
plot(sim)
abline(v = period, col = 'red')
```
```{r}
plot(periodogram(sim)[[1]],
     periodogram(sim)[[2]], type = 'l',
     xlim = c(0,0.5))
```


```{r}
mean(rle(as.vector(sign(diff(sim))))$lengths) * 2
1/best1
1/best2
```
It appears that the observed period matches reasonably well with $\frac{1}{f_2}$ and not with $\frac{1}{f_1}$

\newpage
\subsection{c}
$$X''_t = 0.9X''_{t-5} + Z_t$$
```{r}
library(astsa)
freq = arma.spec(ar = c(0,0,0,0,0.9), var.noise = 1, type = "line")
best1 = 0
best2 = 0.2
best3 = 0.4
```
The frequencies that appear to be dominant are $f_1 \in [0, 0.05]$, $f_2 \in [0.19, 0.21]$, and $f_3 \in [0.39, 0.41]$. The peak occured at 0, 0.2, and 0.4.

```{r}
set.seed(1)
sim = arima.sim(n = 50, list(ar = c(0,0,0,0,0.9)))
period = cumsum(rle(as.vector(sign(diff(sim))))$lengths) + 1
plot(sim)
abline(v = period, col = 'red')
```
```{r}
plot(periodogram(sim)[[1]],
     periodogram(sim)[[2]], type = 'l',
     xlim = c(0,0.5))
```


```{r}
mean(rle(as.vector(sign(diff(sim))))$lengths) * 2
1/best1
1/best2
1/best3
```
It appears that the observed period matches reasonably well with $\frac{1}{f_3}$ and not with $\frac{1}{f_2}$ or $\frac{1}{f_1}$.

\newpage
\section{2}
$$(1-0.9B^3)X_t = Z_t, Z_t \sim WN(1)$$
\subsection{a} 
Compute transfer, power transfer functions, and spectral density $f_X(\lambda)$ associated with AR polynomial $(1-0.9B^3)$.
$$A(\lambda) = \sum_ja_je^{-2\pi ij\lambda}, -1/2\leq \lambda \leq 1/2$$
Where $|A(\lambda)|^2$ is the power transfer function. Thus with $a_0 = 1$ and $a_3 = -0.9$ then,
$$= 1 - 0.9e^{-6\pi i\lambda} = \phi(e^{-2\pi i\lambda})$$
Now, the spectral density is given by,
$$f_X(\lambda) = \sigma_Z^2\frac{|\theta(e^{-2\pi i\lambda})|^2}{|\phi(e^{-2\pi i\lambda})|^2}, -1/2\leq \lambda \leq 1/2$$
With $\theta(z) = 1, \phi(z) = 1-0.9z^3$,
$$= \frac{\sigma_Z^2}{|1 - 0.9*(e^{-2\pi i\lambda})^3|^2}$$
$$= \frac{\sigma_Z^2}{|1 - 0.9*(e^{-6\pi i\lambda})|^2}$$
$$=\frac{\sigma_Z^2}{|1 - 0.9\cos(6\pi\lambda) - 0.9i\sin(6\pi\lambda)|^2}$$
$$= \frac{\sigma_Z^2}{(1 - 0.9\cos(6\pi\lambda))^2 - 0.9^2i^2\sin^2(6\pi\lambda)}$$
$$=\frac{\sigma_Z^2}{1 - 2*0.9\cos(6\pi\lambda) + 0.9^2\cos^2(6\pi\lambda) + 0.9^2\sin^2(6\pi\lambda)}$$
$$=\frac{\sigma_Z^2}{1 - 2*0.9\cos(6\pi\lambda) + 0.9^2(\cos^2(6\pi\lambda)+ \sin^2(6\pi\lambda))}$$
Using $\cos^2(\theta) + \sin^2(\theta) = 1$ then,
$$=\frac{\sigma_Z^2}{1 + 0.9^2 - 2*0.9\cos(6\pi\lambda)}$$
Thus the power transfer function is,
$$|A(\lambda)|^2 = 1 + 0.9^2 - 2*0.9\cos(6\pi\lambda)$$

\newpage
\subsection{b}

Plotting the theoretical spectral density gives,
```{r}
specden <- function(x){1/(1 + 0.9^2 - 2*0.9*cos(6*pi*x))}
grid = seq(from = 0, to = 1/2, length.out = 1000)
grid2 = seq(from = 0.2, to = 1/2, length.out = 1000)
plot(grid, specden(grid), type = 'l')
best1 = grid[which.max(specden(grid))]
best2 = grid2[which.max(specden(grid2))]
```
Based off this plot, I believe the series will oscillate. Since the frequency that contributes the most is `r best2`, I believe it will have period 3.

\newpage
\subsection{c}
```{r}
set.seed(1)
sim = arima.sim(n = 50, list(ar = c(0,0,0.9)))
period = cumsum(rle(as.vector(sign(diff(sim))))$lengths) + 1
plot(sim)
abline(v = period, col = 'red')
```
```{r}
plot(periodogram(sim)[[1]],
     periodogram(sim)[[2]], type = 'l',
     xlim = c(0,0.5))
```

```{r}
mean(rle(as.vector(sign(diff(sim))))$lengths) * 2
1/best1
1/best2
```
Since the period roughly matches the theoretical prediction, the two simulations are consistent.

\newpage
\subsection{d}
Now consider the linear filter with weights $a_1 = a_0 = a_1 = \frac{1}{3}; a_j = 0$ otherwise. 
$$A(\lambda) = \frac{1}{2q + 1}(2\frac{\sin(\pi(q+1)\lambda)}{\sin(\pi\lambda)}\cos(\pi q\lambda)-1)$$
and with $q=1$, 
$$= \frac{1}{3}(2\frac{\sin(2\pi\lambda)}{\sin(\pi\lambda)}\cos(\pi\lambda)-1)$$
Then, the power transfer function would be,
$$|A(\lambda)|^2 = \frac{1}{9}(2\frac{\sin(2\pi\lambda)}{\sin(\pi\lambda)}\cos(\pi\lambda)-1)^2$$
Thus, the spectral density for $Y$ would be,
$$f_Y(\lambda) = |A(\lambda)|^2f_X(\lambda)$$
$$= \frac{1}{9}(2\frac{\sin(2\pi\lambda)}{\sin(\pi\lambda)}\cos(\pi\lambda)-1)^2 * \frac{\sigma_Z^2}{1 + 0.9^2 - 2*0.9\cos(6\pi\lambda)}$$

\newpage
\subsection{e}
```{r}
Xspecden <- function(x){1/(1 + 0.9^2 - 2*0.9*cos(6*pi*x))}
Yspecden <- function(y){abs((1/9)*(2*(sin(2*pi*y)/sin(pi*y))*cos(pi*y)-1))^2}
grid = seq(from = 0, to = 1/2, length.out = 1000)
par(mfrow = c(2,1))
plot(grid, Xspecden(grid), type = 'l')
plot(grid, Yspecden(grid)*Xspecden(grid), type = 'l')
```
Looking at the plot, the frequency at 1/3 has been smoothed. Thus, I would expect the series to oscillate with a period larger than 3 after the smoothing.

\newpage
\subsection{f}
```{r}
set.seed(1)
sim = arima.sim(n = 50, list(ar = c(0,0,0.9)))
sim.filter = filter(sim, sides=2, filter=rep(1/3,3))
period = cumsum(rle(as.vector(sign(diff(sim.filter))))$lengths) + 1
plot(sim.filter)
abline(v = period, col = 'red')
```
```{r}
par(mfrow = c(2,1))
plot(periodogram(sim)[[1]],
     periodogram(sim)[[2]], type = 'l',
     xlim = c(0,0.5))
plot(periodogram(sim.filter[!is.na(sim.filter)])[[1]],
     periodogram(sim.filter[!is.na(sim.filter)])[[2]], type = 'l',
     xlim = c(0,0.5))
```
```{r}
mean(rle(as.vector(sign(diff(sim.filter))))$lengths) * 2
```
The simulation confirms the theory as the average period is now larger than before, thus indicating less weight from the frequency of 1/3.

\newpage
\section{3}
For $x_0,...,x_{n-1}$ and $y_0,...,y_{n-1}$
$$z_t = \sum_{k = 0}^{n-1}x_{t-k}y_k$$
With $x_{-m} = x_{n-m}$. The $j$th coefficient of a DFT is given by,
$$b_j = \sum_{t=0}^{n-1}x_te^{-\frac{2\pi ijt}{n}}, j = 0,...,n-1$$
Thus, 
$$b_j^X = \sum_{t=0}^{n-1}x_te^{-\frac{2\pi ijt}{n}}, j = 0,...,n-1$$
and,
$$b_j^Y = \sum_{t=0}^{n-1}y_te^{-\frac{2\pi ijt}{n}}, j = 0,...,n-1$$
Now consider the product,
$$b_j^Xb_j^Y =  \sum_{t=0}^{n-1}x_te^{-\frac{2\pi ijt}{n}} \cdot \sum_{t=0}^{n-1}y_te^{-\frac{2\pi ijt}{n}}, j = 0,...,n-1$$
$$= \sum_{k = 0}^{n-1}\sum_{t = 0}^{n-1} x_te^{-\frac{2\pi ijt}{n}}y_ke^{-\frac{2\pi ijk}{n}}$$
$$= \sum_{k = 0}^{n-1}\sum_{t = 0}^{n-1}x_ty_ke^{\frac{-2\pi ij}{n}(k + t)}$$
Using the IFT, $x_{t} = \frac{1}{n}\sum_{j = 0}^{n-1}b_je^{\frac{2\pi i j(t)}{n}}$
$$= \frac{1}{n}\sum_{k = 0}^{n-1}\sum_{t = 0}^{n-1}\sum_{j = 0}^{n-1}b_jy_ke^{\frac{-2\pi ij}{n}(k + t)}e^{\frac{2\pi i jt}{n}}$$
$$= \frac{1}{n}\sum_{k = 0}^{n-1}\sum_{t = 0}^{n-1}\sum_{j = 0}^{n-1}b_jy_ke^{\frac{-2\pi ij}{n}(k)}$$
$$=  \frac{1}{n}\sum_{t = 0}^{n-1}\sum_{k = 0}^{n-1}\sum_{j = 0}^{n-1}b_jy_ke^{\frac{2\pi i j(-k)}{n}}$$
$$= \frac{1}{n}\sum_{t = 0}^{n-1}\sum_{k = 0}^{n-1}\sum_{j = 0}^{n-1}b_jy_ke^{-\frac{2\pi ijt}{n}}e^{\frac{2\pi i j(t-k)}{n}}$$
$x_{t-k} = \frac{1}{n}\sum_{j = 0}^{n-1}b_je^{\frac{2\pi i j(t-k)}{n}}$,
$$ = \sum_{t = 0}^{n-1}\big(\sum_{k = 0}^{n-1}x_{t-k}y_k\big)e^{-\frac{2\pi ijt}{n}}$$
$$=\sum_{t = 0}^{n-1}z_te^{-\frac{2\pi ijt}{n}}= b_j^Z,  j = 0,...,n-1$$
$$\therefore b_j^Xb_j^Y  = b_j^Z,  j = 0,...,n-1$$
\newpage
\section{4}
Suppose $x_t$ is $h$-cyclic for some $h$. Also that the DFT of $x_0,...,x_{h-1}$ is $\beta_0,...,\beta_{h-1}$. That is,
$$\{\beta_j\}_{j=0}^{j=h-1} = \{\sum_{t = 0}^{h-1}x_te^{-\frac{2\pi ijt}{n}}\}_{j=0}^{j=h-1}$$
For $n = kh$ data points instead of $h$, note that because $x_{t+h} = x_t \forall t$, then for $\forall j$, 
$$b_j = \sum_{t = 0}^{n-1}x_te^{-\frac{2\pi ijt}{n}} = x_1*e^{-\frac{2\pi ij(1)}{n}} + ... + x_{n-1}*e^{-\frac{2\pi ij(n-1)}{n}}$$
$$= x_1(e^{-\frac{2\pi ij(1)}{n}} + ... + e^{-\frac{2\pi ij(n-h)}{n}}) + ... + x_{h-1}(e^{-\frac{2\pi ij(h-1)}{n}} + ...+ e^{-\frac{2\pi ij(n-1)}{n}})$$
$$= x_1(e^{-\frac{2\pi ij(1)}{n}} + e^{-\frac{2\pi ij(h)}{n}} + ... + e^{-\frac{2\pi ij(n-h)}{n}})+ ... + x_{h-1}(e^{-\frac{2\pi ij(h-1)}{n}} + ...+ e^{-\frac{2\pi ij(n-1)}{n}})$$
Now note that the data can be written in $k$ groups of $h$ observations each,
$$\{t\} = \{0,...,h-1\}\bigcup\{h,...,2h-1\}\bigcup...\bigcup\{(k-1)h = n-h,...,kh-1 = n-1\}$$
So we can rewrite the sums as,
$$= \sum_{t = 0}^{h-1}x_t\sum_{c = 0}^{k-1}e^{(-2\pi ij\frac{t + c\cdot h}{n})}$$
Factoring out the part that does not depend on $c$,
$$= \sum_{t = 0}^{h-1}x_t\cdot e^{-2\pi ij\frac{t}{n}}\sum_{c = 0}^{k-1}(e^{(-2\pi ij\frac{h}{n})})^c$$
Now recognize that for $j = mk$ for some $m\in \mathbb N^+$ and using that $\frac{h}{n} = \frac{1}{k}$,
$$(e^{(-2\pi ij\frac{h}{n})})^c = (e^{(\frac{-2\pi imk}{k})})^c = (e^{(-2\pi im)})^c = 1$$
And since there will be $k$ of these terms of $1$, we can break the sum into two cases,
$$= \sum_{t = 0}^{h-1}x_t\cdot e^{-2\pi ij\frac{t}{n}}(k + \sum_{c = 0, j \neq mk}^{k-1}(e^{(\frac{-2\pi ij}{k})})^c)$$
Now using the geometric formula, $\sum_{i = 0}^{k-1}z^i = \frac{1-z^k}{1-z}$, on the second case,
$$= \sum_{t = 0}^{h-1}x_t\cdot e^{-2\pi ij\frac{t}{n}} (k +\frac{1-(e^{(\frac{-2\pi ij}{k})})^k}{1-e^{(\frac{-2\pi ij}{k})}})$$
$$= \sum_{t = 0}^{h-1}x_t\cdot e^{-2\pi ij\frac{t}{n}}(k + \frac{1-e^{(-2\pi ij)}}{1-e^{(-2\pi ij\frac{1}{k})}})$$
But $\frac{1-e^{(-2\pi ij)}}{1-e^{(-2\pi ij\frac{1}{k})}} = 0$ because the numerator is $0$ and the denominator is nonzero. Thus,
$$= k\sum_{t = 0}^{h-1}x_t\cdot e^{-2\pi ij\frac{t}{n}}$$
$$= k\beta_j, \forall j$$
Thus, if we consider $n = kh$ then $b_j = b_{mk} = k\beta_m$ for $0\leq m \leq h-1$ and $b_j = 0$ otherwise.





